---
title: "Executive Summary: Exploring Small Business Loan Approval Likelihood"
subtitle: |
  | Final Project 
  | Data Science 2 with R (STAT 301-2)
author: "Nikole Montero Cervantes"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon=false}

## Github Repo Link

[https://github.com/stat301-2-2024-winter/final-project-2-Nikole26.git](https://github.com/stat301-2-2024-winter/final-project-2-Nikole26.git)

:::

```{r}
#| label: loading-packages-and-data
#| echo: false
library(tidyverse)
library(here)
```

# Introduction

The project objective is to develop a predictive model to determine the likelihood of small businesses or start-ups having their loan applications approved by the U.S. Small Business Administration (SBA). This predictive model focuses on classifying loan applications into two categories: 'Approved' and 'Rejected,' utilizing the dependent variable loan status, which indicates if the loan is either 'Charged Off' or 'Paid in Full'. The dataset was sourced from Kaggle, and this report presents key insights, findings, and future research recommendations.

# Main Insights and Findings

## Type of prediction problem

This project focus on a classification problem since the project aim to predict whether the loan is Charged Off or Paid in Full.

## Models Assesment

```{r}
#| label: fig-table-confusion-matrix
#| fig-cap: Confusion Matrix
load(here("results/conf_mat.rda"))
print(conf_mat$table)
```


```{r}
#| label: fig-best-model-accuracy
#| fig-cap: Best Model Accuracy
load(here("results/acc_rf.rda"))
acc_rf
```

```{r}
#| label: fig-table-roc
#| fig-cap: ROC (Receiver Operating Characteristic Curve) 
load(here("results/roc_curve.rda"))
roc_curve
```

### Recipe Performance Comparison

## Best Model - Radom Forest

# Opportunity for future research
